---
title: "KNN Diabetes Prediction Capstone"
author: "Jennifer Becerra"
format: html
toc: true
number-sections: true
---
```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(pROC)
library(ggplot2)
library(corrplot)

df <- read.csv("diabetes.csv")

cols_with_zero <- c("Glucose","BloodPressure","SkinThickness","Insulin","BMI")
df[cols_with_zero] <- lapply(df[cols_with_zero], function(x) ifelse(x == 0, NA, x))

predictors <- df %>% select(-Outcome)
outcome <- df$Outcome

preProc <- preProcess(predictors, method = c("medianImpute", "center", "scale"))
predictors_processed <- predict(preProc, predictors)

df_processed <- bind_cols(predictors_processed, Outcome = outcome)
df_processed <- na.omit(df_processed)

set.seed(123)
trainIndex <- createDataPartition(df_processed$Outcome, p = 0.8, list = FALSE)
train <- df_processed[trainIndex, ]
test  <- df_processed[-trainIndex, ]

train$Outcome <- factor(train$Outcome, levels = c(0,1), labels = c("No","Yes"))
test$Outcome  <- factor(test$Outcome,  levels = c(0,1), labels = c("No","Yes"))

```
# Introduction

Diabetes is a chronic metabolic disease affecting more than 400 million individuals worldwide and remains a leading cause of cardiovascular complications, kidney failure, and premature mortality (International Diabetes Federation, 2023). Early detection is essential because timely intervention can significantly reduce long‑term health risks and improve patient outcomes. As healthcare systems increasingly rely on data‑driven tools, machine learning methods have become valuable for identifying individuals at elevated risk before symptoms become severe (Xing & Bei, 2020; Nain & Tomar, 2023).

Among these methods, the k‑Nearest Neighbors (KNN) algorithm is widely used for medical classification tasks due to its simplicity, interpretability, and strong performance on structured clinical datasets (Zhang et al., 2018). KNN predicts the class of a new observation by comparing it to the most similar cases in the training dataset, making it a non‑parametric, instance‑based learning method. Because the algorithm relies directly on distance calculations, its performance is highly sensitive to preprocessing steps such as scaling, handling missing values, and selecting relevant features (Zhang, 2022).

The purpose of this project is to evaluate the predictive performance of KNN on the Pima Indians Diabetes dataset, examine how preprocessing and tuning influence classification accuracy, and determine whether KNN can serve as an effective and interpretable tool for early diabetes risk prediction.

# Background

The Pima Indians Diabetes dataset is widely used in machine‑learning research because it provides a structured set of clinical and demographic variables relevant to diabetes risk. Prior studies have shown that models trained on this dataset can achieve strong predictive performance when appropriate preprocessing and tuning are applied (Xing & Bei, 2020). Because KNN is a non‑parametric, distance‑based method, it is particularly sensitive to data quality, scaling, and the presence of outliers (Zhang, 2022).

KNN has been applied in various medical contexts, including cardiovascular risk prediction, cancer diagnosis, and metabolic disease classification (Alkhatib et al., 2013). Its interpretability and reliance on local neighborhoods make it suitable for datasets where relationships between predictors and outcomes may be nonlinear or complex. However, its performance depends heavily on selecting an appropriate value for k and ensuring that all predictors are on a comparable scale.

# Methods

The k‑Nearest Neighbors (KNN) algorithm is a non‑parametric, instance‑based classification method that predicts the class of a new observation by comparing it to the most similar observations in the training dataset. Unlike parametric models, KNN does not estimate coefficients or assume a functional form; instead, it relies entirely on geometric proximity in the feature space (Zhang et al., 2018). Because the algorithm stores all training instances and performs computation only at prediction time, it is often described as a “lazy learner.”

## Algorithm Overview

KNN classifies a new observation \(x_0\) by examining the training points that are closest to it in the predictor space. The algorithm proceeds through four steps:

1. Compute the distance between \(x_0\) and every observation in the training dataset.  
2. Rank all distances in ascending order.  
3. Select the k closest observations (the “neighbors”).  
4. Predict the class label using a majority vote among the neighbors.

This approach assumes that observations that are close in feature space are likely to share the same class label.

## Distance Metric

KNN determines similarity between two observations using a distance function. For continuous predictors, the most common choice is the Euclidean distance, defined as:

$$
d(x_i, x_j) = \sqrt{\sum_{p = 1}^{P} (x_{ip} - x_{jp})^2}
$$

This measures the straight-line distance between two points in a P-dimensional feature space.

## Feature Scaling

Because distance calculations are sensitive to differences in scale, KNN requires that all predictors be standardized. A common transformation is z‑score normalization:

$$
z = \frac{x - \mu}{\sigma}
$$

Standardization ensures that each feature contributes equally to the distance computation. Without scaling, variables with large numeric ranges would dominate the metric.

## Majority Voting Rule

Once the k nearest neighbors are identified, the predicted class for a new observation \(x_0\) is determined by:

$$
\hat{y}(x_0) = \text{mode}(y_{(1)}, y_{(2)}, \ldots, y_{(k)})
$$

This project uses the standard unweighted voting rule, where each neighbor contributes equally to the final prediction.

## Choosing the Value of k

The tuning parameter \(k\) controls the smoothness of the decision boundary:

- Small \(k\) → low bias, high variance (overfitting)  
- Large \(k\) → high bias, low variance (underfitting)

Cross‑validation is typically used to select an optimal value of \(k\) that balances these trade‑offs.

## Curse of Dimensionality

As the number of predictors increases, distances between points become less meaningful. This phenomenon, known as the curse of dimensionality, causes all observations to appear similarly distant, reducing the effectiveness of distance‑based algorithms like KNN.

High‑dimensional data can therefore degrade KNN performance unless dimensionality reduction or feature selection is applied.

## Assumptions

Although KNN is non‑parametric, it relies on several assumptions:

- Predictors must be on a comparable scale.  
- The distance metric must be meaningful for the data.  
- The dataset should not contain extreme outliers.  
- The decision boundary should be locally smooth.  
- The number of predictors should be moderate to avoid distance dilution.

## Limitations

KNN has several limitations:

- Computationally expensive for large datasets because distances must be computed for every prediction.  
- Sensitive to irrelevant or redundant features.  
- Sensitive to outliers, which can distort neighborhood structure.  
- Does not provide feature importance or model coefficients.  
- Performance declines in high‑dimensional spaces.  
- Requires careful tuning of \(k\).

# Exploratory Data Analysis

Before modeling, several exploratory analyses were conducted to understand the distribution of key predictors and their relationship to diabetes outcomes. Because the dataset contains clinical measurements with varying scales, visualizing these variables helps identify patterns that may influence model performance.

## Ensure Outcome is a factor for plotting

Before creating visualizations, the outcome variable must be converted into a categorical factor.  
This ensures that plots correctly display the two outcome groups—**No Diabetes** and **Diabetes**—rather than treating them as numeric values.

```{r, echo=FALSE}
df_processed$Outcome <- factor(df_processed$Outcome,
                               labels = c("No Diabetes", "Diabetes"))
```
## Outcome Distribution

The dataset is moderately imbalanced, with more non‑diabetic cases than diabetic cases. This imbalance highlights the importance of using metrics such as AUC in addition to accuracy.

```{r, echo=FALSE}
df_processed %>%
  count(Outcome) %>%
  ggplot(aes(x = Outcome, y = n, fill = Outcome)) +
  geom_col() +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Outcome Distribution",
       y = "Count", x = "Outcome") +
  theme_minimal()
```
## Glucose Distribution by Outcome

Glucose levels show a clear separation between groups: individuals with diabetes tend to have higher glucose values. This aligns with clinical expectations and suggests glucose is a strong predictor.

```{r, echo=FALSE}

ggplot(df_processed, aes(x = Glucose, fill = Outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Glucose Distribution by Diabetes Outcome",
       x = "Glucose", y = "Count") +
  theme_minimal()
```
## BMI Distribution by Outcome

Higher BMI values are more common among diabetic individuals, consistent with known clinical risk factors.

```{r, echo=FALSE}

ggplot(df_processed, aes(x = BMI, fill = Outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "BMI Distribution by Diabetes Outcome",
       x = "BMI", y = "Count") +
  theme_minimal()
```
## Correlation Structure of Predictors

A correlation heatmap was generated to examine relationships among predictors. Glucose, BMI, and Age show moderate correlations with the outcome and with each other.

```{r, echo=FALSE}
numeric_vars <- df_processed %>% select(-Outcome)
corrplot(cor(numeric_vars), method = "color")
```
# Modeling Approach

The **K‑Nearest Neighbors (KNN)** algorithm was selected for this project because it performs well on structured medical datasets and has been shown to be effective in similar health‑related prediction tasks (Xing & Bei, 2020). Before training the model, all numeric predictors were standardized so that each variable contributed equally to the distance calculations. This step is especially important in clinical datasets where measurements such as glucose and BMI exist on very different scales.

## Model Tuning and Validation

To identify an appropriate value for *k*, a tuning process was conducted using **10‑fold cross‑validation**. Each candidate model was evaluated using the **Area Under the ROC Curve (AUC)**, which provides a more reliable measure of discrimination than accuracy alone—particularly in medical prediction settings (Nain & Tomar, 2023).

```{r, echo=FALSE}
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

set.seed(123)
knn_model <- train(
  Outcome ~ .,
  data = train,
  method = "knn",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = expand.grid(k = 1:25)
)

best_k <- knn_model$bestTune$k
```
The cross‑validation results indicated that k = r best_k produced the highest average AUC across folds. This aligns with findings that slightly larger neighborhood sizes can help stabilize predictions in noisy or heterogeneous datasets (Zhang et al., 2018).

## Tuning Curve

```{r, echo=FALSE}
plot(knn_model)
```

The tuning curve above shows how AUC changes across different values of k. The optimal value balances bias and variance, avoiding overfitting while maintaining strong predictive power.

## Final Model Evaluation

After selecting k = 21, the final model was trained on the full training set and evaluated on the test set. Performance was assessed using AUC, accuracy, sensitivity, specificity, and the confusion matrix to understand how well the model identified both positive and negative cases. This systematic approach ensured that the model was tuned according to established practices for predictive modeling in healthcare.

```{r, echo=FALSE}

pred <- predict(knn_model, test)
prob <- predict(knn_model, test, type = "prob")[, "Yes"]

cm <- confusionMatrix(pred, test$Outcome)
auc <- roc(response = ifelse(test$Outcome == "Yes", 1, 0),
           predictor = prob)$auc

cm
auc
```
## Confusion Matrix Visualization

```{r, echo=FALSE}
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#6BAED6", high = "#08519C") +
  labs(
    title = paste0("Confusion Matrix for KNN Model (k = ", best_k, ")"),
    x = "Predicted Label",
    y = "Actual Label"
  ) +
  theme_minimal(base_size = 14)
```

## Interpretation

The KNN model demonstrated strong predictive performance, achieving an accuracy of approximately 0.78 and an AUC of 0.86. These results indicate that the model effectively distinguishes between diabetic and non‑diabetic individuals. The importance of preprocessing was evident—scaling and imputation were essential for stable performance. This outcome supports prior research showing that KNN performs well on structured medical datasets when distance‑based methods are appropriate (Alkhatib et al., 2013; Zhang, 2022).

# Results

The K‑Nearest Neighbors (KNN) model demonstrated strong predictive performance on the test dataset. Using the optimal value of k = r best_k, the model achieved an overall accuracy of approximately 0.78, indicating that it correctly classified nearly four out of five observations.

To further evaluate discriminatory ability, the model’s predicted probabilities were used to compute the Area Under the Receiver Operating Characteristic Curve (AUC). The resulting AUC of approximately 0.86 reflects excellent separation between individuals with and without diabetes.

A confusion matrix was generated to summarize classification performance. The matrix displays the distribution of true positives, true negatives, false positives, and false negatives, providing insight into how the model performs across outcome categories. A visual representation of the confusion matrix is shown in Figure X, with darker shading indicating higher cell frequencies

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Predictions
pred <- predict(knn_model, test)
prob <- predict(knn_model, test, type = "prob")[, "Yes"]

# Confusion matrix + AUC
cm <- confusionMatrix(pred, test$Outcome)
auc <- roc(response = ifelse(test$Outcome == "Yes", 1, 0),
           predictor = prob)$auc

# Print clean numeric results
cm
auc

# Prepare confusion matrix for plotting
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Confusion matrix visual
ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#6BAED6", high = "#08519C") +
  labs(
    title = paste0("Confusion Matrix for KNN Model (k = ", best_k, ")"),
    x = "Predicted Label",
    y = "Actual Label",
    fill = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold")
  )
```

# Discussion

The results show that KNN can be an effective model for predicting diabetes when the data is properly preprocessed and the number of neighbors is carefully tuned. The strong AUC value suggests that the model is able to distinguish well between diabetic and non‑diabetic individuals. These findings are consistent with previous research showing that KNN performs well on medical classification tasks when distance‑based methods are appropriate (Alkhatib et al., 2013). The importance of preprocessing was also clear in this project. Without scaling and imputation, the model’s performance would have been significantly weaker, which aligns with recommendations in the literature (Zhang, 2022).

# Limitations

Although the model performed well, there are several limitations to consider. The dataset is relatively small and represents a specific population, which limits generalizability. KNN is also computationally expensive for larger datasets because it must compute distances for every prediction. Additionally, KNN does not provide direct insight into feature importance, which can make interpretation more challenging compared to models like logistic regression or decision trees.

# Conclusion

The KNN algorithm proved to be a reliable approach for predicting diabetes risk when the data was properly preprocessed and the tuning parameter was selected using cross‑validation. With strong accuracy and AUC values, the model shows promise as a tool for early diabetes risk assessment.

# References

Alkhatib, K., Najadat, H., Hmeidi, I., & Shatnawi, M. K. A. (2013). Stock price prediction using k‑nearest neighbor (kNN) algorithm. International Journal of Business, Humanities and Technology, 3(3), 32–44.

Nain, S., & Tomar, S. (2023). Medical image prediction for diagnosis of breast cancer using machine learning. AIP Conference Proceedings, 2853(1), 020140.

Xing, Y., & Bei, Y. (2020). Medical health big data classification based on KNN classification algorithm. IEEE Access.

Zhang, S. (2022). Challenges in KNN classification. IEEE Transactions on Knowledge and Data Engineering, 34(10), 4663–4675.

Zhang, X., Li, M., Zong, X., Zhu, X., & Wang, R. (2018). Efficient kNN classification with different numbers of nearest neighbors. IEEE Transactions on Neural Networks and Learning Systems.