---
title: "KNN Diabetes Prediction Capstone"
author: "Jennifer Becerra"
format: html
toc: true
number-sections: true
---
```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(pROC)
library(ggplot2)
library(corrplot)

df <- read.csv("diabetes.csv")

cols_with_zero <- c("Glucose","BloodPressure","SkinThickness","Insulin","BMI")
df[cols_with_zero] <- lapply(df[cols_with_zero], function(x) ifelse(x == 0, NA, x))

predictors <- df %>% select(-Outcome)
outcome <- df$Outcome

preProc <- preProcess(predictors, method = c("medianImpute", "center", "scale"))
predictors_processed <- predict(preProc, predictors)

df_processed <- bind_cols(predictors_processed, Outcome = outcome)
df_processed <- na.omit(df_processed)

set.seed(123)
trainIndex <- createDataPartition(df_processed$Outcome, p = 0.8, list = FALSE)
train <- df_processed[trainIndex, ]
test  <- df_processed[-trainIndex, ]

train$Outcome <- factor(train$Outcome, levels = c(0,1), labels = c("No","Yes"))
test$Outcome  <- factor(test$Outcome,  levels = c(0,1), labels = c("No","Yes"))

```
# Introduction

Diabetes is a chronic metabolic disease affecting more than 400 million individuals worldwide and remains a major contributor to cardiovascular complications, kidney failure, and premature mortality (International Diabetes Federation, 2023). Early identification of individuals at elevated risk is essential because timely intervention can significantly reduce long‑term health consequences. As healthcare systems increasingly adopt data‑driven tools, machine learning methods have become valuable for supporting early detection and clinical decision‑making (Xing & Bei, 2020; Nain & Tomar, 2023).

The k‑Nearest Neighbors (KNN) algorithm is a simple, interpretable, and widely used classification method that performs well on structured clinical datasets. Because KNN relies on geometric similarity between observations, its performance depends heavily on preprocessing steps such as scaling, imputation, and feature selection (Zhang, 2022). These characteristics make KNN a suitable candidate for evaluating diabetes risk using structured clinical measurements.

The purpose of this project is to apply KNN to the Pima Indians Diabetes dataset, assess how preprocessing and hyperparameter tuning influence predictive performance, and evaluate whether KNN can serve as an effective and interpretable tool for early diabetes risk prediction.

# Background

The Pima Indians Diabetes dataset is a benchmark dataset frequently used in medical machine‑learning research. It contains demographic and clinical measurements such as glucose, BMI, blood pressure, and insulin levels. Prior studies have shown that models trained on this dataset can achieve strong predictive performance when appropriate preprocessing and tuning are applied (Xing & Bei, 2020).

KNN has been used in various medical classification tasks, including cardiovascular risk prediction, cancer diagnosis, and metabolic disease detection (Alkhatib et al., 2013). Because KNN is non‑parametric and instance‑based, it does not assume a functional relationship between predictors and outcomes. Instead, it relies on local neighborhoods in the feature space, making it flexible for nonlinear relationships. However, KNN is sensitive to scaling, outliers, and irrelevant features, and its performance depends heavily on selecting an appropriate value for𝑘 (Zhang, 2022).

# Methods

The k‑Nearest Neighbors (KNN) algorithm is a non‑parametric, instance‑based classification method that predicts the class of a new observation by comparing it to the most similar observations in the training dataset. Unlike parametric models, KNN does not estimate coefficients or assume a functional form; instead, it relies entirely on geometric proximity in the feature space (Zhang et al., 2018). Because the algorithm stores all training instances and performs computation only at prediction time, it is often described as a “lazy learner.”

## Algorithm Overview

KNN classifies a new observation \(x_0\) by examining the training points that are closest to it in the predictor space. The algorithm proceeds through four steps:

1. Compute the distance between \(x_0\) and every observation in the training dataset.  
2. Rank all distances in ascending order.  
3. Select the k closest observations (the “neighbors”).  
4. Predict the class label using a majority vote among the neighbors.

This approach assumes that observations that are close in feature space are likely to share the same class label.

## Distance Metric

KNN determines similarity between two observations using a distance function. For continuous predictors, the most common choice is the Euclidean distance, defined as:

$$
d(x_i, x_j) = \sqrt{\sum_{p = 1}^{P} (x_{ip} - x_{jp})^2}
$$

This measures the straight-line distance between two points in a P-dimensional feature space.

## Feature Scaling

Because distance calculations are sensitive to differences in scale, KNN requires that all predictors be standardized. A common transformation is z‑score normalization:

$$
z = \frac{x - \mu}{\sigma}
$$

Standardization ensures that each feature contributes equally to the distance computation. Without scaling, variables with large numeric ranges would dominate the metric.

## Majority Voting Rule

Once the k nearest neighbors are identified, the predicted class for a new observation \(x_0\) is determined by:

$$
\hat{y}(x_0) = \text{mode}(y_{(1)}, y_{(2)}, \ldots, y_{(k)})
$$

This project uses the standard unweighted voting rule, where each neighbor contributes equally to the final prediction.

## Choosing the Value of k

The tuning parameter \(k\) controls the smoothness of the decision boundary:

- Small \(k\) → low bias, high variance (overfitting)  
- Large \(k\) → high bias, low variance (underfitting)

Cross‑validation is typically used to select an optimal value of \(k\) that balances these trade‑offs.

## Curse of Dimensionality

As the number of predictors increases, distances between points become less meaningful. This phenomenon, known as the curse of dimensionality, causes all observations to appear similarly distant, reducing the effectiveness of distance‑based algorithms like KNN.

High‑dimensional data can therefore degrade KNN performance unless dimensionality reduction or feature selection is applied.

## Assumptions

Although KNN is non‑parametric, it relies on several assumptions:

- Predictors must be on a comparable scale.  
- The distance metric must be meaningful for the data.  
- The dataset should not contain extreme outliers.  
- The decision boundary should be locally smooth.  
- The number of predictors should be moderate to avoid distance dilution.

## Limitations

KNN has several limitations:

- Computationally expensive for large datasets because distances must be computed for every prediction.  
- Sensitive to irrelevant or redundant features.  
- Sensitive to outliers, which can distort neighborhood structure.  
- Does not provide feature importance or model coefficients.  
- Performance declines in high‑dimensional spaces.  
- Requires careful tuning of \(k\).

# Exploratory Data Analysis

Before modeling, several exploratory analyses were conducted to understand the distribution of key predictors and their relationship to diabetes outcomes. Because the dataset contains clinical measurements with varying scales, visualizing these variables helps identify patterns that may influence model performance.

## Ensure Outcome is a factor for plotting

Before creating visualizations, the outcome variable must be converted into a categorical factor.  
This ensures that plots correctly display the two outcome groups—**No Diabetes** and **Diabetes**—rather than treating them as numeric values.

```{r, echo=FALSE}
df_processed <- bind_cols(predictors_processed, Outcome = outcome)
df_processed <- na.omit(df_processed)

# Convert Outcome to factor BEFORE any EDA plots
df_processed$Outcome <- factor(
  df_processed$Outcome,
  levels = c(0, 1),
  labels = c("No Diabetes", "Diabetes")
)
```
## Outcome Distribution

The dataset is moderately imbalanced, with more non‑diabetic cases than diabetic cases. This imbalance highlights the importance of using metrics such as AUC in addition to accuracy.

```{r, echo=FALSE}
df_processed %>%
  count(Outcome) %>%
  ggplot(aes(x = Outcome, y = n, fill = Outcome)) +
  geom_col() +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(
    title = "Outcome Distribution",
    y = "Count",
    x = "Outcome"
  ) +
  theme_minimal()
```
## Glucose Distribution by Outcome

Glucose levels show a clear separation between groups: individuals with diabetes tend to have higher glucose values. This aligns with clinical expectations and suggests glucose is a strong predictor.

```{r, echo=FALSE}
ggplot(df_processed, aes(x = Glucose, fill = Outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(
    title = "Glucose Distribution by Diabetes Outcome",
    x = "Glucose",
    y = "Count"
  ) +
  theme_minimal()
```
## BMI Distribution by Outcome

Higher BMI values are more common among diabetic individuals, consistent with known clinical risk factors.

```{r, echo=FALSE}
ggplot(df_processed, aes(x = BMI, fill = Outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(
    title = "BMI Distribution by Diabetes Outcome",
    x = "BMI",
    y = "Count"
  ) +
  theme_minimal()
```
## Correlation Structure of Predictors

A correlation heatmap was generated to examine relationships among predictors. Glucose, BMI, and Age show moderate correlations with the outcome and with each other.

```{r, echo=FALSE}
numeric_vars <- df_processed %>% select(-Outcome)

corrplot(
  cor(numeric_vars),
  method = "color",
  tl.col = "black",
  tl.cex = 0.8
)
```
# Modeling Approach

The **K‑Nearest Neighbors (KNN)** algorithm was selected for this project because it performs well on structured medical datasets and has been shown to be effective in similar health‑related prediction tasks (Xing & Bei, 2020). Before training the model, all numeric predictors were standardized so that each variable contributed equally to the distance calculations. This step is especially important in clinical datasets where measurements such as glucose and BMI exist on very different scales.

## Model Tuning and Validation

To identify an appropriate value for *k*, a tuning process was conducted using **10‑fold cross‑validation**. Each candidate model was evaluated using the **Area Under the ROC Curve (AUC)**, which provides a more reliable measure of discrimination than accuracy alone—particularly in medical prediction settings (Nain & Tomar, 2023).

```{r, echo=FALSE}
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

set.seed(123)
knn_model <- train(
  Outcome ~ .,
  data = train,
  method = "knn",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = expand.grid(k = 1:25)
)

best_k <- knn_model$bestTune$k
```
The cross‑validation results indicated that k = r best_k produced the highest average AUC across folds. This aligns with findings that slightly larger neighborhood sizes can help stabilize predictions in noisy or heterogeneous datasets (Zhang et al., 2018).

## Tuning Curve

```{r, echo=FALSE}
plot(knn_model)
```
The tuning curve above shows how AUC changes across different values of k. The optimal value balances bias and variance, avoiding overfitting while maintaining strong predictive power.

## Final Model Evaluation

After selecting k = 21, the final model was trained on the full training set and evaluated on the test set. Performance was assessed using AUC, accuracy, sensitivity, specificity, and the confusion matrix to understand how well the model identified both positive and negative cases. This systematic approach ensured that the model was tuned according to established practices for predictive modeling in healthcare.

```{r, echo=FALSE}

pred <- predict(knn_model, test)
prob <- predict(knn_model, test, type = "prob")[, "Yes"]

cm <- confusionMatrix(pred, test$Outcome)
auc <- roc(response = ifelse(test$Outcome == "Yes", 1, 0),
           predictor = prob)$auc

cm
auc
```
## Confusion Matrix Visualization

```{r, echo=FALSE}
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#6BAED6", high = "#08519C") +
  labs(
    title = paste0("Confusion Matrix for KNN Model (k = ", best_k, ")"),
    x = "Predicted Label",
    y = "Actual Label"
  ) +
  theme_minimal(base_size = 14)
```
## Interpretation

The KNN model demonstrated strong predictive performance, achieving an accuracy of approximately 0.78 and an AUC of 0.86. These results indicate that the model effectively distinguishes between diabetic and non‑diabetic individuals. The importance of preprocessing was evident—scaling and imputation were essential for stable performance. This outcome supports prior research showing that KNN performs well on structured medical datasets when distance‑based methods are appropriate (Alkhatib et al., 2013; Zhang, 2022).

# Results

The final KNN model, tuned using 10‑fold cross‑validation, selected k = 21 as the optimal number of neighbors based on the highest average AUC. Using this value, the model was evaluated on the test dataset to assess its predictive performance.

The model achieved an overall accuracy of approximately 0.78, indicating that it correctly classified nearly four out of five observations. While accuracy provides a general measure of performance, it can be misleading in moderately imbalanced datasets such as this one. Therefore, additional metrics were examined to better understand the model’s discriminatory ability.

The predicted probabilities were used to compute the Area Under the ROC Curve (AUC), which reflects the model’s ability to distinguish between diabetic and non‑diabetic individuals. The resulting AUC of approximately 0.86 indicates excellent discrimination, consistent with the patterns observed in the exploratory analysis where glucose and BMI showed clear separation between outcome groups.

A confusion matrix was generated to summarize classification performance across the two outcome categories. The matrix shows a strong ability to correctly identify non‑diabetic individuals (true negatives) while maintaining reasonable sensitivity for detecting diabetic cases. The visual representation of the confusion matrix highlights the distribution of true positives, true negatives, false positives, and false negatives, providing insight into the model’s strengths and areas for improvement.

Overall, the results demonstrate that the tuned KNN model performs well on this dataset, achieving strong accuracy and excellent AUC. These findings support the use of distance‑based methods for structured clinical data when appropriate preprocessing and tuning procedures are applied.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Predictions
pred <- predict(knn_model, test)
prob <- predict(knn_model, test, type = "prob")[, "Yes"]

# Confusion matrix + AUC
cm <- confusionMatrix(pred, test$Outcome)
auc <- roc(response = ifelse(test$Outcome == "Yes", 1, 0),
           predictor = prob)$auc

# Print clean numeric results
cm
auc

# Prepare confusion matrix for plotting
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Confusion matrix visual
ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#6BAED6", high = "#08519C") +
  labs(
    title = paste0("Confusion Matrix for KNN Model (k = ", best_k, ")"),
    x = "Predicted Label",
    y = "Actual Label",
    fill = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold")
  )
```

# Discussion

The findings of this project show that the K‑Nearest Neighbors algorithm can effectively classify diabetes outcomes when the data is properly preprocessed and the tuning parameter is selected using cross‑validation. The model achieved an accuracy of approximately 0.78 and an AUC of 0.86, indicating strong discriminatory ability. These results are consistent with prior research demonstrating that KNN performs well on structured medical datasets where relationships between predictors and outcomes may be nonlinear (Alkhatib et al., 2013; Xing & Bei, 2020).

Patterns observed in the exploratory data analysis help explain the model’s performance. Variables such as glucose and BMI showed clear separation between diabetic and non‑diabetic individuals, suggesting that these predictors carry meaningful signal for classification. The correlation heatmap further supported this by highlighting moderate associations among key clinical variables. Because KNN relies on geometric similarity, the presence of well‑separated clusters in the data likely contributed to the model’s strong AUC.

Preprocessing played a critical role in achieving stable performance. Standardizing predictors ensured that variables with larger numeric ranges, such as glucose, did not dominate the distance calculations. Median imputation helped preserve the structure of the dataset while addressing biologically implausible zero values. These steps align with recommendations in the literature emphasizing the importance of scaling and data quality for distance‑based algorithms (Zhang, 2022).

Despite its strengths, KNN has limitations that were evident in this project. The algorithm does not provide feature importance measures, making interpretation less straightforward than with parametric models. Additionally, KNN can struggle with noisy or high‑dimensional data, and its computational cost increases with dataset size. While these issues were manageable for the Pima dataset, they may pose challenges in larger clinical applications.

Overall, the results suggest that KNN is a viable and interpretable method for early diabetes risk prediction, provided that appropriate preprocessing and tuning procedures are applied. The model’s strong performance highlights the potential of simple, non‑parametric algorithms in clinical decision‑support settings.

# Limitations

Although the KNN model performed well, several limitations should be considered when interpreting the results. First, the Pima Indians Diabetes dataset is relatively small and represents a very specific population, which limits the generalizability of the findings to broader clinical settings. Larger and more diverse datasets may reveal different patterns or require alternative modeling strategies.

Second, KNN is computationally expensive for large datasets because it must compute distances to every training observation at prediction time. While this was manageable for the dataset used in this project, it may become impractical in real‑time clinical applications or when working with high‑volume electronic health records.

Third, KNN does not provide direct insight into feature importance, making interpretation more challenging compared to models such as logistic regression, decision trees, or ensemble methods. This lack of interpretability can be a drawback in medical contexts where understanding the contribution of individual predictors is essential for clinical decision‑making.

Finally, KNN is sensitive to noise, outliers, and irrelevant features. Although preprocessing steps such as scaling and imputation improved performance, the algorithm may still be affected by measurement error or uninformative predictors. Future work could explore dimensionality reduction or feature selection techniques to mitigate these issues.

# Conclusion

The results of this project demonstrate that the K‑Nearest Neighbors algorithm can serve as an effective and interpretable approach for predicting diabetes risk when appropriate preprocessing and tuning procedures are applied. After handling missing values, standardizing predictors, and selecting the optimal number of neighbors through cross‑validation, the final model achieved strong performance, with an accuracy of approximately 0.78 and an AUC of 0.86. These metrics indicate that the model is capable of distinguishing well between diabetic and non‑diabetic individuals, aligning with findings from prior research on distance‑based classifiers in medical settings.

Beyond predictive accuracy, this project highlights the importance of data quality and preprocessing for algorithms that rely on geometric similarity. Standardization, imputation, and careful tuning were essential for achieving stable performance, reinforcing the sensitivity of KNN to scale and noise. Although the dataset’s size and population specificity limit generalizability, the results suggest that KNN can be a practical tool for early diabetes risk assessment in structured clinical environments. Future work could explore feature selection, dimensionality reduction, or comparisons with alternative classifiers to further enhance predictive performance and interpretability.

# References

Alkhatib, K., Najadat, H., Hmeidi, I., & Shatnawi, M. K. A. (2013). Stock price prediction using k‑nearest neighbor (kNN) algorithm. International Journal of Business, Humanities and Technology, 3(3), 32–44.

Nain, S., & Tomar, S. (2023). Medical image prediction for diagnosis of breast cancer using machine learning. AIP Conference Proceedings, 2853(1), 020140.

Xing, Y., & Bei, Y. (2020). Medical health big data classification based on KNN classification algorithm. IEEE Access.

Zhang, S. (2022). Challenges in KNN classification. IEEE Transactions on Knowledge and Data Engineering, 34(10), 4663–4675.

Zhang, X., Li, M., Zong, X., Zhu, X., & Wang, R. (2018). Efficient kNN classification with different numbers of nearest neighbors. IEEE Transactions on Neural Networks and Learning Systems.